Child survival for mothers: mortality change and related measures

### 2020-12-28 Old/paper.Rmd
code para reproducible

---
title: 'Child survival for mothers: mortality change and related measures'
author: "Iván Williams and Diego Alburez-Gutierrez"
date: "May 2020"
output:
  pdf_document: default
  html_document:
    fig_caption: yes
bibliography: bib.bibtex
---

@@

The question of child survival, particularly the survival of daughters for mothers, sits at the very center of demographic theory.
This is evidenced by the omnipresence of the 'net reproductive rate' in accounts of human and non-human populations. 
Historical demographers draw liberally on assumptions about kin availability and individual's exposure to offspring mortality to explain fertility decisions, especially around the demographic transition. 
These assumptions are often untested given the data sparsity [@livi_bacci_concise_1997;@volk_infant_2013]. 
Economic theories of fertility that consider children as 'investment' for old-age support also rely on notions of child survival [@preston1978effects].

Kin count estimation has a long and distinguished pedigree in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations across demographic regimes. To the best of our knowledge, @brass_derivation_1953 first proposed a formula to represent child survival over age. The notion of estimating the expected number of living daughters in a stable population was generalized by @goodman1974 for other kin relations (granddaughters, cousins, etc.). The "counting method" approach was further popularized in @KeyfitzMath. @bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'. Recently @Caswell2019 reformulated these relationships in a matrix framework, extending the potential results to variance estimation, multistate models, and others.

Child survival - and it's counterpart, child death - gain new significance in the wake of the fertility transition.
Depressed fertility means that fewer children are tasked with providing key emotional, social, and financial transfers to aging parents for ever increasing periods of time.
With increasing periods of generational overlap, individuals find themselves 'sandwiched' between aging parents and young children requiring their simultaneous attention and care [@daatland_norwegian_2010].
In the context of global population aging, elderly parents without access to formal social security and pension systems are particularly reliant on these transfers to make ends meet [@smith-greenaway_maternal_2020].  
Losing an only child may become a more common experience in the context of global fertility decline, something particularly worrying for parents suddenly rendered childless.
Given the known psychological, health, and social consequences of child loss for bereaved parents and families [@Hendrickson2009;@lee_death_2014], it is surprising that the demographic processes that shape parental bereavement remain very poorly understood.
Recent development have underscored the need to understand how sudden changes in mortality affect the availability of kin. 
In the context of the global pandemic caused by the Covid-19 disease, elderly people depend on support provided by their relatives, especially during periods of lockdown and in the absence of governmental support mechanisms.<!-- Maybe here are repeated --> Elderly parents may be at a higher risk of losing the key support provided by their adult children given the known age-gradient in the Covid-19 case fatality rate.

How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother? How can we characterize the timing of offspring mortality as experienced by a mother from a formal demographic perspective? How this would work in an heterogeneity population?
In the remainder of the paper, we aim to formalize the relationship population-level changes in mortality rates and perceived change in the lived experience of child death from the perspective of a mother. Our approach is relevant for understanding the impact of overall and age-specific excess mortality on the resilience or otherwise of kinship networks for the vulnerable elderly population.
We consider the case when these changes are additive and mutliplicative. In this paper, we uncover formal relationships to answer these questions and provide numerical applications using data from countries in Latin America during the second half of the twentieth century. See the [Applications](#Applications) section for more details on the source, calculation procedures and main numerical results. We used period measures, in an stable context, so the objetive is give moment measures. A more sophisticated model using the appropriate age-specific cohort rates in the subscripts has been proposed by @alburez_gutierrez_womens_2019.

@@

That an equal change in mortality could happens at all ages is unlikelihood. This measure could be much related a change in first ages (child and youth) or adolescent ones. If the change only happens in the age range $\left[0;1\right)$, infant mortality, we can inspect the effect in child survival $CS_{a}^{\delta_0}$ starting with splitting the integral between those daughters who were exposed all the range and those who do not (now generalizing for $\alpha$ positive and $a>1$ for simplicity):

@@

$$\frac{dCS^{\delta_0}}{d\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}$$
If we name $CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}$, we can express the first term in the right as $CS_a - CS_{a-1,a}$, and recognizing that last term could be expressed as  $\kappa_{a-1,a} CS_{a-1,a}$, we get:

$$ \frac{dCS^{\delta_0}}{d\delta_0} = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})$$
Where the factor $1-(a-\kappa_{a-1,a})$, which lies between 0 and 1, is equal to the average portion of time spent with living daughters durings age *a*. Assuming uniform distribution of those births, discretizing the change and dividing by $CS_a$ we get the final expression for change in infant mortality:

@@

This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (cohorts aged more than 1), except for "half" the portion $\frac{CS_{a-1,a}}{CS_a}$ on how important was this las age in total "successful" experience (in terms of alive descendants). If $a$ is near $\beta$ then $\frac{\Delta CS_{a}^{\delta_0}}{CS_{a}} \approx -\Delta\delta$.

@@

The last integral is $CS_{a-t}$. Dividing by $CS_a$ and multiplying by $\delta$ get one useful expression. This is the negative of cumulative hazard $H_a$ but considering a positive factor $\frac{CS_{a-t}}{CS_a}$ that takes in account the relative amount of surviving descendants that would be lost at each child age at risk, that gives more weight to first ages because all the parity was exposed in that age.

@@

Grasping this relationship intuitively is more difficult given the interaction of birth and mortality rates (@KeyfitzMath). 
The factor $\frac{CS_{a-t}}{CS_a}$  has a S-shape because $l_x$ curvature and fertility accumulation. Taking the difference between consecutive ages $a$ and $a+1$, we have $CS_{a-1}=\int_{0}^{a-1} {m_{x} l_{a-1} \, dx}$, $CS_{a}=\int_{0}^{a-1} {m_{x} l_{a} \, dx}+CS_{a-1,a}$. In the last one, a second term is added because of additional born daughters after $a-1$, and in the first term the existing one are survived at on $\frac{l_{a-x}}{l_{a-1-x}}$. In advanced ages when fertility is complete, the difference is only explained by the surviving part, that is why this factor is not a strictly increasing function but for mother ages close to $\beta$ (last theoretical fecundity age). A numerical evaluation of both analytical expressions (absolute an relative) and its factor $\frac{CS_{a-t}}{CS_a}$ are done in [Applications](#Applications) section.

>>>>

---
title: 'Child survival for mothers: mortality change and related measures'
author: "Iván Williams and Diego Alburez-Gutierrez"
date: "May 2020"
output:
  pdf_document: 
    keep_tex: yes
  html_document:
    fig_caption: yes
bibliography: bib.bibtex
---

@@

The question of child survival for mothers sits at the very center of demographic theory.
Offspring survival is usually studied alongside fertility decline in the context of the demographic transition, but it also matters for post-transitional societies. 
Lower fertility and higer life expectancy means that fewer children are expected to provide key emotional, social, and financial transfers to aging parents for longer periods of time.
With increasing periods of generational overlap, individuals find themselves 'sandwiched' between aging parents and young children requiring their simultaneous attention and care [@daatland_norwegian_2010].
In the context of global population aging, elderly parents without access to formal social security and pension systems are particularly reliant on these transfers [@smith-greenaway_maternal_2020].
Given the known psychological, health, and social consequences of child loss for bereaved parents and families [@Hendrickson2009;@lee_death_2014], it is surprising that the demographic processes that shape parental bereavement remain very poorly understood.

The Covid-19 pandemic has underscored the need to understand how sudden changes in mortality affect the availability of kin. 
The elderly depend on kin support, especially during periods of lockdown and in the absence of governmental support mechanisms.
Elderly mothers may be at a higher risk of losing the key support provided by their adult children given the known age-gradient in the Covid-19 case fatality rate.
How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother? 
How do these processes play out in a heterogeneous population?

In this paper, we aim to formalize the relationship between population-level changes in mortality rates (additive or multiplicative) and changes in the lived experience of kin death.
<!-- Ivan: si decimos esto, quiza deberia ser el enfoque principale del paper tener age-specific changes?  -->
This is key for understanding the impact of age-specific excess mortality on the resilience or otherwise of kinship networks for the vulnerable elderly population.
This paper uncovers formal relationships and provides numerical applications using data from Latin America in the second half of the twentieth century using period measures in a stable context.^[See the [Applications](#Applications) section for more details on the source, calculation procedures and main numerical results.]
We continue a long tradition in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations, @brass_derivation_1953 on child survival, @goodman1974 on kin survival in stable populations, and recently, @caswell_formal_2019 kinship matrices.

<!-- The question of child survival, particularly the survival of daughters for mothers, sits at the very center of demographic theory. -->
<!-- This is evidenced by the omnipresence of the 'net reproductive rate' in accounts of human and non-human populations.  -->
<!-- Historical demographers draw liberally on assumptions about kin availability and individual's exposure to offspring mortality to explain fertility decisions, especially around the demographic transition.  -->
<!-- These assumptions are often untested given the data sparsity [@livi_bacci_concise_1997;@volk_infant_2013].  -->
<!-- Economic theories of fertility that consider children as 'investment' for old-age support also rely on notions of child survival [@preston1978effects]. -->

<!-- Kin count estimation has a long and distinguished pedigree in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations across demographic regimes. To the best of our knowledge, @brass_derivation_1953 first proposed a formula to represent child survival over age. The notion of estimating the expected number of living daughters in a stable population was generalized by @goodman1974 for other kin relations (granddaughters, cousins, etc.). The "counting method" approach was further popularized in @KeyfitzMath. @bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'. Recently @Caswell2019 reformulated these relationships in a matrix framework, extending the potential results to variance estimation, multistate models, and others. -->

<!-- Child survival - and it's counterpart, child death - gain new significance in the wake of the fertility transition. -->
<!-- Depressed fertility means that fewer children are tasked with providing key emotional, social, and financial transfers to aging parents for ever increasing periods of time. -->
<!-- With increasing periods of generational overlap, individuals find themselves 'sandwiched' between aging parents and young children requiring their simultaneous attention and care [@daatland_norwegian_2010]. -->
<!-- In the context of global population aging, elderly parents without access to formal social security and pension systems are particularly reliant on these transfers to make ends meet [@smith-greenaway_maternal_2020].   -->
<!-- Losing an only child may become a more common experience in the context of global fertility decline, something particularly worrying for parents suddenly rendered childless. -->
<!-- Given the known psychological, health, and social consequences of child loss for bereaved parents and families [@Hendrickson2009;@lee_death_2014], it is surprising that the demographic processes that shape parental bereavement remain very poorly understood. -->
<!-- Recent development have underscored the need to understand how sudden changes in mortality affect the availability of kin.  -->
<!-- In the context of the global pandemic caused by the Covid-19 disease, elderly people depend on support provided by their relatives, especially during periods of lockdown and in the absence of governmental support mechanisms. -->
<!-- Elderly parents may be at a higher risk of losing the key support provided by their adult children given the known age-gradient in the Covid-19 case fatality rate.  -->

<!-- How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother? How can we characterize the timing of offspring mortality as experienced by a mother from a formal demographic perspective? How this would work in an heterogeneity population? -->
<!-- In the remainder of the paper, we aim to formalize the relationship population-level changes in mortality rates and perceived change in the lived experience of child death from the perspective of a mother. Our approach is relevant for understanding the impact of overall and age-specific excess mortality on the resilience or otherwise of kinship networks for the vulnerable elderly population. -->
<!-- We consider the case when these changes are additive and mutliplicative. In this paper, we uncover formal relationships to answer these questions and provide numerical applications using data from countries in Latin America during the second half of the twentieth century. See the [Applications](#Applications) section for more details on the source, calculation procedures and main numerical results. We used period measures, in an stable context, so the objetive is give moment measures. A more sophisticated model using the appropriate age-specific cohort rates in the subscripts has been proposed by @alburez_gutierrez_womens_2019. -->

@@

However, it is unlikely for mortality to change at the same rate at all ages. 
In the course of the demographic transition, for example, we would expect larger changes in young-age mortality.
<!-- This measure could be much related a change in first ages (child and youth) or adolescent ones.  -->
Consider a change in infant mortality in the age range $\left[0;1\right)$. We can inspect the effect in child survival $CS_{a}^{\delta_0}$ by splitting the integral between those daughters who were exposed to the mortality change and those who were not (now generalizing for $\alpha$ positive and $a>1$ for simplicity):

@@

$$\frac{dCS^{\delta_0}}{d\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}$$
Given that $CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}$, we can express the first term in the right as $CS_a - CS_{a-1,a}$. 
The righ-most term could also be expressed as  $\kappa_{a-1,a} CS_{a-1,a}$, so that:

$$ \frac{dCS^{\delta_0}}{d\delta_0} = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})$$
The factor $1-(a-\kappa_{a-1,a})$, which lies between 0 and 1, is equal to the average portion of time spent with living daughters durings age *a*. Assuming a uniform distribution of those births, discretizing the change and dividing by $CS_a$ we get the final expression for change in infant mortality:

@@

<!-- Este parrafo no esta muy claro... -->
This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (cohorts aged more than 1), except for "half" the portion $\frac{CS_{a-1,a}}{CS_a}$ on how important was this las age in total "successful" experience (in terms of alive descendants). If $a$ is near $\beta$ then $\frac{\Delta CS_{a}^{\delta_0}}{CS_{a}} \approx -\Delta\delta$.

@@

Considering that the last integral is equal to $CS_{a-t}$, we divide by $CS_a$ and multiply by $\delta$ to obtain:

@@

This is the negative of cumulative hazard $H_a$ but considering a positive factor $\frac{CS_{a-t}}{CS_a}$ that takes in account the relative amount of surviving descendants that would be lost at each child age at risk, that gives more weight to first ages because all the parity was exposed in that age.

Grasping this relationship intuitively is more difficult given the interaction of birth and mortality rates (@KeyfitzMath). 
The factor $\frac{CS_{a-t}}{CS_a}$  has a S-shape because $l_x$ curvature and fertility accumulation. Taking the difference between consecutive ages $a$ and $a+1$, we have $CS_{a-1}=\int_{0}^{a-1} {m_{x} l_{a-1} \, dx}$, $CS_{a}=\int_{0}^{a-1} {m_{x} l_{a} \, dx}+CS_{a-1,a}$. In the last one, a second term is added because of additional born daughters after $a-1$, and in the first term the existing one are survived at on $\frac{l_{a-x}}{l_{a-1-x}}$. In advanced ages when fertility is complete, the difference is only explained by the surviving part, that is why this factor is not a strictly increasing function but for mother ages close to $\beta$ (last theoretical fecundity age). We provide a numerical evaluation of both analytical expressions (absolute an relative) in the [Applications](#Applications) section.
